---
title: "Lab 5"
author: "Ayesha Gamage-(ayega981)/Muditha Cherangani(mudch175)"
date: "2023-12-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Hypothesis testing

### 1.Scatterplot of Y versus X
```{r, echo=FALSE}
library("boot")
########Read data from csv file
setwd("D:/MSC/SEMESTER1/732A90  Computational statistics/Labs/submites")
data<-read.delim("lottery.csv",sep=";")

##### Plot data
plot(data$Day_of_year, data$Draft_No, main="Scatterplot (Y=Draft No) versus  (X=Day of year)",
     xlab="Day of Year", ylab="Draft No", pch=19)
```

The data points are spread out in this graph. This means there is no data trend; thus, there is no pattern visible in the scatterplot.

### 2. Fit a curve to the data
```{r, echo=FALSE}
#linear regression model
data1 <- data[-4]
lm_model<-lm(Draft_No~.,data1)
# summary(lm_model)

##### Plot data
plot(data$Day_of_year, data$Draft_No, main="Scatterplot (Y=Draft No) versus  (X=Day of year)",
     xlab="Day of Year", ylab="Draft No", pch=19)

pred.int.lin<-predict(lm_model,data,interval="conf",level=0.95)
lines(data$Day_of_year,pred.int.lin[,1],col="yellow",lwd=2)

lm.p3<-lm(formula=Draft_No~poly(Day_of_year,3),data)
pred.int.p3<-predict(lm.p3,data,interval="conf",level=0.95)
lines(data$Day_of_year,pred.int.p3[,1],col="green",lwd=4)

lm.phigh<-lm(formula=Draft_No~poly(Day_of_year,20),data)
pred.int.phigh<-predict(lm.phigh,data,interval="conf",level=0.95)
lines(data$Day_of_year,pred.int.phigh[,1],col="red",lwd=2)


loess_model <- loess(Draft_No~Day_of_year,data)
predicted_values <- predict(loess_model, newdata = data)

lines(data$Day_of_year, predicted_values, col = "blue",lwd=2)
legend("topright", legend = c("lm-default","lm-p3", "lm-p20","loess"), col = c("yellow","green","red", "blue"),lty = 1)
```
The yellow, red, and green line represents the prediction of the linear regression model.The blue line represents the prediction of the loess model. From the point of view of the lines, there is a clear relationship between parameters.There are correlation between X and Y. However, these curves are not enough to identify which parameters are significantly responsible for non–randomness. When consider the order ,order is significant for the curve because the highest order given the approximately equel curve to loess model's curve.

### 3.Check if the lottery is random
The statistic consider,
$$
S=\sum_{i = 1}^{n}|\hat{Y}_{i}-\overline{Y}|
$$
```{r, echo=FALSE}
observed_s <- sum(abs(predict(loess_model) - mean(data$Draft_No)))
cat("S = ",observed_s)
```
Estimate S’s distribution through a non–parametric bootstrap, B = 2000 
```{r, echo=FALSE,warning=FALSE}
set.seed(1234)
stat1<-function(data,indices){
  train_data<-as.data.frame(data[indices,])
  res<-lm(Draft_No~Day_of_year,train_data)
  res$coefficients[2]
}

res<-boot(data,stat1,R=2000)
ci <- boot.ci(res)
print(ci)
```
```{r, echo=FALSE}
# Extract p-values from the confidence interval
cat("P-values :",pt(ci$t0,length(data$Day_of_year)-1))
plot(res)
```

The p–value of the observed value of S,
```{r, echo=FALSE}
set.seed(1234)
# Function to calculate the statistic S
statistic_S <- function(data, indices, loess_model) {
  observed_values <- data$Draft_No
  predicted_values <- predict(loess_model)
  
  # Calculate the statistic S for the current bootstrap sample
  S <- sum(abs(predicted_values[indices] - mean(observed_values)))
  
  return(S)
}

# Using boot function  
  loess_model <- loess(Draft_No~Day_of_year,data)
  res <- boot(data, statistic_S, R = 2000, loess_model = loess_model)

# Calculate the observed S statistic
observed_S <- statistic_S(data, 1:nrow(data), loess_model)

# Calculate the p-value
p_value <- mean(res$t >= observed_S)

# Display the p-value
cat("p-value:", p_value, "\n")

```
According to BOOTSTRAP CONFIDENCE INTERVAL CALCULATIONS, confidence intervals levels 95%. Computed p value is $0.4955 > 0.05 $ , reject $H_{0}$ ,then lottery wasn't random.


### 4.Based on the test statistic S, implemented a function that tests the hypothesis,
$$
    H_{0}: Lottery  \ is \ random versus  \\
    H_{1}: Lottery \ is \ non–random.
$$
The function return the value of S and its p–value, based on 2000 bootstrap samples.
```{r, echo=TRUE}
  set.seed(1234)
# Function to test the hypothesis
test_hypothesis <- function(data, loess_model, B = 2000) {
  # Using boot function
  res <- boot(data, statistic_S, R = B, loess_model = loess_model)
  
  # Calculate the observed S statistic
  observed_S <- statistic_S(data, 1:nrow(data), loess_model)
  
  # Calculate the p-value
  p_value <- mean(res$t >= observed_S)
  
  # Return results
  result <- list(observed_S = observed_S, p_value = p_value)
  return(result)
}
  # Example call
  result <- test_hypothesis(data, loess_model)

  # Display results
  cat("Observed S:", result$observed_S, "\n")
  cat("p-value:", result$p_value, "\n")

```

### 5.Create a dataset of the same dimensions as the original data.Here I used consecutive dates and blocks of consecutive dates for create data.


```{r, echo=FALSE}
# Function to create a dataset with assigned end numbers for consecutive dates
create_dataset_consecutive <- function(n, k) {
  # Generate dates from 1 to n
  dates <- 1:n
  
  # Create a dataset with Date and LotteryNumber columns
  dataset <- data.frame(Day_of_year = dates, Draft_No = NA)
  
  # Choose k consecutive dates and assign them end numbers
  start_index <- sample(1:(n - k + 1), 1)
  end_dates <- start_index:(start_index + k - 1)
  dataset$Draft_No[end_dates] <- n + 1  # Assign end numbers
  
  # Assign random numbers to the remaining dates
  remaining_dates <- dataset$Draft_No[is.na(dataset$Draft_No)]
  remaining_dates <- sample(1:(n - k), length(remaining_dates), replace = FALSE)
  dataset$Draft_No[is.na(dataset$Draft_No)] <- remaining_dates
  
  # Return the dataset
  return(dataset)
}
```


```{r, echo=FALSE}
create_dataset_blocks <- function(n, k) {
  # Generate dates from 1 to n
  dates <- 1:n
  
  # Create a dataset with Date and LotteryNumber columns
  dataset <- data.frame(Day_of_year = dates, Draft_No = NA)
  
  # Ensure k is at least 3
  if (k < 3) {
    stop("k should be at least 3 for blocks of consecutive dates.")
  }
  # Calculate the number of blocks (bk/3c)
  num_blocks <- floor(k/3)
  block_start <- sample(seq(1, n - 2, by = 3), num_blocks, replace = FALSE)
   for (j in block_start) {
    #cat("block_start",j,"block_end",j+2,"\n")
    dataset$Draft_No[j:(j+2)] <- n + 1  # Assign end numbers
  }
  remaining_dates <- dataset$Draft_No[is.na(dataset$Draft_No)]
  #cat("1",remaining_dates,"\n")
  # If k is not divisible by 3, some blocks can be of length bk/3c + 1
  if (k %% 3 != 0) {
    remaining_block <- sample(which(is.na(dataset$Draft_No)), k-(num_blocks*3))
    #cat("rest",k-(num_blocks*3),remaining_block,"\n")
    dataset$Draft_No[remaining_block] <- n + 1  # Assign end number
  }
 
  remaining_dates <- dataset$Draft_No[is.na(dataset$Draft_No)]
  #cat("2",length(remaining_dates),"\n")
  if(length(remaining_dates)>0){
    remaining_dates <- sample(1:(n - k), length(remaining_dates), replace = FALSE)
    dataset$Draft_No[is.na(dataset$Draft_No)] <- remaining_dates
  }
 
  
  # Return the dataset
  return(dataset)
}


```


```{r, echo=FALSE}
n <- 366
k <- 50  # You can adjust this value

# Create the dataset with consecutive dates
generated_dataset_consecutive <- create_dataset_consecutive(n, k)

# Fit the loess model to the consecutive dataset
loess_consecutive <- loess(Draft_No ~ Day_of_year, data = generated_dataset_consecutive)
# Create the dataset with blocks of consecutive dates
generated_dataset_blocks <- create_dataset_blocks(n, k)
t<-which(is.na(generated_dataset_blocks$Draft_No))
generated_dataset_blocks$Draft_No[is.na(generated_dataset_blocks$Draft_No)]
# Fit the loess model to the block dataset
loess_blocks <- loess(Draft_No ~ Day_of_year, data = generated_dataset_blocks)

# Test hypothesis for consecutive dataset
result_consecutive <- test_hypothesis(generated_dataset_consecutive, loess_consecutive)

# Test hypothesis for block dataset
result_blocks <- test_hypothesis(generated_dataset_blocks, loess_blocks)

# Display results
cat("Results for Consecutive Dataset:\n")
cat("Observed S:", result_consecutive$observed_S, "\n")
cat("p-value:", result_consecutive$p_value, "\n")
cat("Accept H0:", result_consecutive$p_value < 0.05, "\n\n")

cat("Results for Block Dataset:\n")
cat("Observed S:", result_blocks$observed_S, "\n")
cat("p-value:", result_blocks$p_value, "\n")
cat("Accept H0:", result_blocks$p_value < 0.05, "\n")

```
```{r, echo=FALSE}
k = 10
Consecutive_result <- list()
Block_result <- list()
for(k in 3:(n-1)){
  # Create the dataset with consecutive dates
  generated_dataset_consecutive <- create_dataset_consecutive(n, k)
  
  # Fit the loess model to the consecutive dataset
  loess_consecutive <- loess(Draft_No ~ Day_of_year, data = generated_dataset_consecutive)
  
  
  # Create the dataset with blocks of consecutive dates
  generated_dataset_blocks <- create_dataset_blocks(n, k)
  # Fit the loess model to the block dataset
  loess_blocks <- loess(Draft_No ~ Day_of_year, data = generated_dataset_blocks)

  # Test hypothesis for consecutive dataset
  result_consecutive <- test_hypothesis(generated_dataset_consecutive, loess_consecutive)
  
  # Test hypothesis for block dataset
  result_blocks <- test_hypothesis(generated_dataset_blocks, loess_blocks)
  # cat("K value:", k, "\n")
  if(0.025 < abs(result_consecutive$p_value) || abs(result_consecutive$p_value)  > 0.025){

    Consecutive_result[[length(Consecutive_result)+1]] <- list(Observed_S <-result_consecutive$observed_S,p_value <- result_consecutive$p_value)
    Block_result[[length(Block_result)+1]] <- list(Observed_S <-result_blocks$observed_S,p_value <-result_blocks$p_value)
    
    #here i print only 3 sample for the report.
    if(k==0 || k==5 || k==10){
    cat("Results for Consecutive Dataset:\n")
    cat("Observed S:", result_consecutive$observed_S, "\n")
    cat("p-value:", result_consecutive$p_value, "\n")
    cat("Reject H0:", result_consecutive$p_value < 0.025, "\n\n")

    cat("Results for Block Dataset:\n")
    cat("Observed S:", result_blocks$observed_S, "\n")
    cat("p-value:", result_blocks$p_value, "\n")
    cat("Reject H0:", result_blocks$p_value < 0.025, "\n")
    
    }
  }
}
ggplot(economics, aes(date, unemploy)) + geom_line()
```
In this case, computed p-values are approximately equal, and all p-values are greater than 0.05. Then $H_{0}$ rejected.


